---
title: "Class 17 Questions 13&14"
author: "Dan Vu (PID: A17380158)"
format: pdf
---

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
table(expr$geno)
```
I can infer from the plot created that A/A has a higher relative expression value compared to    G/G. This means that the SNP does affect the expression of ORMDL3. There are 108 samples of A/A, 233 samples of A/G, and 121 samples of G/G.

```{r}
library(ggplot2)
```

>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
ggplot(expr) +
  aes(geno, exp, col = geno) +
  geom_boxplot(notch = TRUE)
```

I can infer from the plot created that A/A has a higher relative expression value compared to    G/G. This means that the SNP does affect the expression of ORMDL3.
