---
title: "Class_05: Data Viz with ggplot"
author: "Dan Vu (PID: A17380158)"
format: pdf
---

Today we are exploring the **ggplot** package and how to make nice figures in R.

There are lots of ways to make figures and plots in R. These include: 

- so called "base" R
- and add on packages like **ggplot2**

Here is a simple "base" R plot.

```{r}
head(cars)
```

We can simply pass it to the `plot()` function. 

```{r}
plot(cars)
```

> Key-point: Base R is quick but not so nice looking in some folks eyes.

Let's see how we can plot this with **ggplot2**...

1st I need to instal this add-on package. For this we use the `install.packages()` function - **WE DO THIS IN THE CONSOLE, NOT our report** This is a one time only deal.

2nd We need to load the package with `library()` function every time we want to use it.

```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot is composed of at least 3 layers:

-**data** (i.e a data.frame with the things you want to plot),
-aesthetics **aes()** that map the columns of data to your plot features (i.e aesthetics)
- geoms like **geom_point()** that sort how the plot appears
```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point()
```

```{r}
hist(cars$speed)
```

> Key point: For simple "canned" graphs base R is quicker and more concise to write the code, but as things get more custom and elaborate then ggplot wins out...

Let's add more layers to our ggplot

Add a line showing the relationship between x and y.
Add a title
Add custom axis labels "Speed (MPH)" and "Distance (ft)" 
Change the theme...

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  labs(title = "Silly plot of Speed vs Stopping Distance") +
  labs(x="Speed (MPH)", y="Distance (ft)")+
  theme_bw()
```

## Going further

Read some gene expression data
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

> Q1. How many genes are in this dataset?

```{r}
nrow(genes)
```


> Q2. How many "up" regulated genes are there?

```{r}
sum(genes$State=="up")
```

A useful function for counting up occurences of things in a vector is the `table()` function.

```{r}
table(genes$State)
```

Make a v1 figure

```{r}
p <- ggplot(genes) +
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point()
```

```{r}
p + 
  scale_color_manual(values=c("blue","gray","red"))+
  labs(x = "Control (No Drug)", y = "Drug Treatment") +
  theme_bw()
```

## More Plotting

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder, 3)
```

> Q4. How many different country values are in this dataset?

```{r}
length(table(gapminder$country))
```

```{r}
unique(gapminder$continent)
```

```{r}
ggplot(gapminder) +
  aes(x=gdpPercap, y=lifeExp, col=continent) +
  geom_point()
```

```{r}
ggplot(gapminder) +
  aes(x=gdpPercap, y=lifeExp, col=continent, label=country) +
  geom_point() +
  geom_text()
```

I can use the **ggrepl** package to make more sensible labels here. 

```{r}
library(ggrepel) 
ggplot(gapminder) +
  aes(x=gdpPercap, y=lifeExp, col=continent, label=country) +
  geom_point() +
  geom_text_repel()
```

```{r}
ggplot(gapminder) +
  aes(x=gdpPercap, y=lifeExp, col=continent, label=continent) +
  geom_point() +
  facet_wrap(~continent)
```

##Summary

ggplot2 offers several advantages over base R plots:

1. **Layered Grammar**: ggplot uses a consistent, layered approach—data, aesthetics, and geometric layers—making complex plots easier to build and customize step by step [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

2. **Publication Quality**: ggplot produces attractive, publication-ready figures with sensible defaults, which are often more visually appealing than base R plots [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

3. **Customization**: While base R gives pixel-level control, ggplot makes it easier to add customizations (like colors, labels, themes) and combine multiple elements in a single plot [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

4. **Consistency**: The same building blocks (data, aesthetics, geoms) apply to all plot types, reducing the need to learn different functions for each plot style as in base R [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

5. **Scalability**: For simple plots, base R is quick, but ggplot is more concise and manageable for complex, multi-layered figures [[1]](https://drive.google.com/file/d/1BYSWJLROqxA1YpuDhJkzUolhiZqiOOKg/view?usp=drivesdk), [[2]](https://drive.google.com/file/d/1tFqKg9_nhVMmKYfiM1CQKDS2PmPwLh8n/view?usp=drivesdk), [[3]](https://drive.google.com/file/d/1Clw2_EJ_hY3USNwObiPnxpIQIfirxfW0/view?usp=drivesdk), [[5]](https://drive.google.com/file/d/15xXaaIcCWOc_x1gJLdySWOd_sfMXTiaw/view?usp=drivesdk), [[4]](https://drive.google.com/file/d/1FDBbIi2Rlw2In9mClB7Mub8oUPgx6y8h/view?usp=drivesdk).

